SFJ: An implementation of Semantic Featherweight Java

### 2020-02-26 conference/paper.tex
now?

Consider the set of polygons, such as triangles, rectangles, rumbles given by a class hierarchy.
We want to define a method \emph{diagonal} that takes a polygon and returns the length of its longest diagonal. This method makes sense only if the polygon passed to it has at least four sides, hence triangles are excluded.
In Java this could be implemented in the following ways:
$$
\begin{array}{l}
\synclass Polygon \ \{ \ldots\}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\syndecl{Other\_Polygons}{Polygon} {
\\
	\qquad \mathbf{real}\ diagonal(Other\_Polygons \ p)\ \{\ldots \}
	\\
	\hspace{3.5cm} \ldots
	\\
}
\\
\syndecl{Rectangle}{Other\_Polygons}{\ldots}
\\
\syndecl{Rumble}{Other\_Polygons} {\ldots}
\end{array}
$$
Or by means of an interface \emph{Diagonal}:
$$
\begin{array}{l}
public \ \syninterface Diagonal\ \{
	\\
	\qquad \mathbf{real}\ diagonal(Polygon \ p);
	\\
\}
\\
\synclass Polygon \ \{\ldots\}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\synidecl{Rectangle}{Polygon} \synimpl Diagonal \ \{\ldots\}
\\
\synidecl{Rumble}{Polygon} \synimpl Diagonal \ \{\ldots\}
\\
\hspace{4.5cm} \ldots
\end{array}
$$

@@

We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle} thus allowing the typechecker to check at compile time the restrictions on types:
$$
\begin{array}{l}
\syndecl{Polygon}{Object} {\ldots}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\syndecl{Rectangle}{Polygon} {\ldots}
\\
\syndecl{Rumble}{Polygon} {\ldots}
\\
\hspace{2.5cm} \ldots
\\
\syndecl{Diagonal}{Object} {
	\\
	\qquad \mathbf{real}\ diagonal(\textit{Polygon} \textbf{ and not }  \textit{Triangle} \ p)\{\ldots\}
	\\
}
\end{array}
$$
We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}
\emph{On multimethods}
Since FJ is a core language, some features of the full Java are removed: overloading methods is one of them.
In our framework, leveraging the expressivity of boolean connectives and semantic subtyping, many of these features, and language constructs, e.g., lists, are restored: overloading methods is one of them. We can thus model \emph{multimethods}, \cite{BC97}, which according to the authors is ``\emph{very clean and easy to understand [...] it would be the best solution for a brand new language}".
As an example Dardha \emph{et al.}  \cite{Dardha2013,Dardha2017}, consider the following class declarations:
$$
\begin{array}{ll}
\syndecl{A}{\mathit{Object}} {
&
\qquad
\syndecl{B}{A} {
\\
	%\qquad\quad \ldots
%&
	%\qquad\qquad\quad \ldots
%\\
	\qquad\quad \mathbf{int}\ \mathit{length}\ (\mathbf{string}\ s)\{\ \ldots\ \}\qquad
&
	\qquad\qquad\quad \mathbf{int}\ \mathit{length}\ (\mathbf{int}\ n)\{\ \ldots\ \}
\\
}
&
\qquad }
\end{array}
$$
Method {\it length} has type $\mathbf{string} \rightarrow \mathbf{int}$ in $A$. However, in $B$ it has type $(\mathbf{string} \rightarrow \mathbf{int}) \synwedge (\mathbf{int} \rightarrow \mathbf{int})$,
which can be simplified as $(\mathbf{string} \synvee \mathbf{int}) \rightarrow \mathbf{int}$.

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
As already discussed in the introduction, the key idea behind semantic subtyping is interpreting types as sets, in particular,  as sets of values of the language at hand. Since we want these types
to be able to be finitely representable, we restrict our types to finite trees whose
leaves are constants with no cycles. For example a recursive type $\alpha = [a : \alpha]$ would
be an infinite tree \textbf{new} $C$(\textbf{new} $C$($\ldots{}$)). Similarly the types $\alpha =
[b: \beta]$, $\beta = [a = \alpha]$ would also be impossible to instantiate.

Therefore, when processing the abstract syntax tree (AST) of a program, we do not allow the type of
the class we are defining to be the type of its own field. Secondly we mark any classes with only
basic types \textit{(bool, int, float ...)} for the types of its fields as \textit{resolved} and otherwise, if it contains a class type, we mark it as \textit{unresolved}. After processing the whole AST, we perform the following algorithm to decide whether the type definitions in the program are valid.
\begin{verbatim}
do
    boolean resolutionOccured = false
    for class that is unresolved:
        boolean resolved = true
        for field that contains a class type:
            if class type is not resolved
                resolved = false
        if resolved is true:
            set class to be resolved
            resolutionOccured = true
while resolutionOccured is true

>>>>

Consider the set of polygons, such as triangles, rectangles, rumbles given by a class hierarchy.
We want to define a method \emph{diagonal} that takes a polygon and returns the length of its longest diagonal. This method makes sense only if the polygon passed to it has at least four sides, hence triangles are excluded.
In Java this could be implemented in the following ways:
$$
\begin{array}{l}
\synclass Polygon \ \{ \ldots\}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\syndecl{Other\_Polygons}{Polygon} {
\\
	\qquad \mathbf{real}\ diagonal(Other\_Polygons \ p)\ \{\ldots \}
	\\
	\hspace{3.5cm} \ldots
	\\
}
\\
\syndecl{Rectangle}{Other\_Polygons}{\ldots}
\\
\syndecl{Rhombus}{Other\_Polygons} {\ldots}
\end{array}
$$
Or by means of an interface \emph{Diagonal}:
$$
\begin{array}{l}
public \ \syninterface Diagonal\ \{
	\\
	\qquad \mathbf{real}\ diagonal(Polygon \ p);
	\\
\}
\\
\synclass Polygon \ \{\ldots\}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\synidecl{Rectangle}{Polygon} \synimpl Diagonal \ \{\ldots\}
\\
\synidecl{Rhombus}{Polygon} \synimpl Diagonal \ \{\ldots\}
\\
\hspace{4.5cm} \ldots
\end{array}
$$

@@

We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle} thus allowing the typechecker to check at compile time the restrictions on types:
$$
\begin{array}{l}
\syndecl{Polygon}{Object} {\ldots}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\syndecl{Rectangle}{Polygon} {\ldots}
\\
\syndecl{Rhombus}{Polygon} {\ldots}
\\
\hspace{2.5cm} \ldots
\\
\syndecl{Diagonal}{Object} {
	\\
	\qquad \mathbf{real}\ diagonal(\textit{Polygon} \textbf{ and not }  \textit{Triangle} \ p)\{\ldots\}
	\\
}
\end{array}
$$
We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}
\emph{On multimethods}
Since FJ is a core language, some features of the full Java are removed: overloading methods is one of them.
In our framework, leveraging the expressivity of boolean connectives and semantic subtyping, many of these features, and language constructs, e.g., lists, are restored: overloading methods is one of them. We can thus model \emph{multimethods}, \cite{BC97}, which according to the authors is ``\emph{very clean and easy to understand [...] it would be the best solution for a brand new language}".
As an example Dardha \emph{et al.}  \cite{Dardha2013,Dardha2017}, consider the following class declarations:
$$
\begin{array}{ll}
\syndecl{A}{\mathit{Object}} {
&
\qquad
\syndecl{B}{A} {
\\
	%\qquad\quad \ldots
%&
	%\qquad\qquad\quad \ldots
%\\
	\qquad\quad \mathbf{int}\ \mathit{length}\ (\mathbf{string}\ s)\{\ \ldots\ \}\qquad
&
	\qquad\qquad\quad \mathbf{int}\ \mathit{length}\ (\mathbf{int}\ n)\{\ \ldots\ \}
\\
}
&
\qquad }
\end{array}
$$
Method {\it length} has type $\mathbf{string} \rightarrow \mathbf{int}$ in $A$. However, in $B$ it has type $(\mathbf{string} \rightarrow \mathbf{int}) \synwedge (\mathbf{int} \rightarrow \mathbf{int})$,
which can be simplified as $(\mathbf{string} \synvee \mathbf{int}) \rightarrow \mathbf{int}$.

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
As already discussed in the introduction, the key idea behind semantic subtyping is interpreting types as sets, in particular,  as sets of values of the language at hand. Since we want these types to be able to be finitely representable, we restrict our types to finite trees whose leaves are constants with no cycles. 
For example a recursive type $\alpha = [a : \alpha]$ would be an infinite tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)). 
Similarly the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ would also be impossible to instantiate.

Therefore, when processing the abstract syntax tree (AST) of a program in SFJ, we do not allow the type of the class we are defining to be a type of its own field. 
Secondly we mark any classes with only basic types \emph{(bool, int, float ...)} for the types of its fields as \emph{resolved} and otherwise, if it contains a class type, we mark it as \emph{unresolved}. 
After processing the whole AST, we perform the following algorithm to decide whether the type definitions in the program are valid.
\begin{verbatim}
do
    boolean resolutionOccured = false
    for class that is unresolved:
        boolean resolved = true
        for field that contains a class type:
            if class type is not resolved
                resolved = false
        if resolved is true:
            set class to be resolved
            resolutionOccured = true
while resolutionOccured is true
